The Name of the Title is Hope

### 2023-09-15 input/2_introduction.tex
pre-submission text

\section{Introduction}
% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
\todo{add something even more general here: e.g. increase human wellbeing by XY}
Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities as well as simplifying skill acquisition~\citep{Kasahara2019-sk, Goto2020-mw}. There are two main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e. exoskeletons, a user's body can be moved by applying forces to the extremities, for example by pulling on the fingertips. Another technology makes the user's extremities move by sending current into their muscle activating nerves using electrical muscle stimulation (EMS). However, augmented users often report dissociative experiences, frequently disrupting their sense of agency~\citep{Gilbert2017-ze, Gilbert2019-uc}.

Having a sense of agency means experiencing being in control of our own voluntary actions, instead of them feeling as though they randomly happen to us. It has frequently been shown that if users have a strong sense of agency, they are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system~\citep{Berberian2012-do, Miller2007-rb}. The key challenge to drive adoption of human \textit{action augmentation} then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.
% if something happens for you in an interaction there is a cost to that:
% - impacts sense of agency which in turn may makes user's engage less -> citation?
% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
% What have others done to address this and what do we propose "instead"
Previously, ~\citet{Kasahara2019-sk} have shown that reaction times when tapping a stimuli on a touchscreen can be reduced by augmenting the user with EMS while maintaining their sense of agency. The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited. However, setting this gain factor only works in stimulus-response scenarios where the user behavior can be predicted with high accuracy. A key question then remains: How can we maintain agency during physical action augmentation without controlling the environment? In other words, how can we design a closed-loop system to deliver an integrated, \textit{natural}, experience for users' augmented actions?
% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.

\missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm. <- better use this for the teaser image. Here better present the main results, i.e. sense of agency scores?}
\input{input/1_teaser_figure}

% 4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this short
In this paper, we present a fast communication channel between a user's brain signals and a physical end effector, here EMS. Crucially, our brain-computer interface controls the user's muscles through EMS at the time of their \textit{intent to act}, as measured through readiness potentials (RP) present in the user's electroencephalogram (EEG). In our user study we applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their sense of agency.
% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.

% \subsection{Preserving Agency using Brain Signals of the Intent to (Inter)act}
% 4. (long) Anwser with teaser image
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.

>>>>

\section{Introduction}
% 1. Describe the Current Situtation
% Functions as a starting point and a common basis. Therefore it primarily contains recognizable and agreed points.
Advances in hardware that augment a user's physical actions have reignited dreams of overcoming human limitations, recovering lost abilities and simplifying skill acquisition~\citep{Goto2020-mw}. However, augmented users often report dissociative experiences, frequently disrupting their sense of agency (SoA)~\citep{Gilbert2017-ze, Gilbert2019-uc}.

% 2. What is the complication, challenge identified
% Spells the reason for acting now. It contains threats / opportunities and the hurdles that need to be overcome.
Having a SoA means experiencing control of our own voluntary actions, instead of them feeling as randomly happening to us. It has frequently been shown that users are more likely to feel engaged and satisfied with an interaction, and are more likely to trust a system the more they experience SoA~\citep{Berberian2012-do, Miller2007-rb}. A key challenge to drive the adoption of human action augmentation then is to design for agency experience, so users feel as though they are in the ``driver's seat'' once again.

% 3. Question
% Asks the question how the hurdles of the Complication can be overcome. How can we prevent the threat or seize the opportunity? Also, what would be the benefits if the complication would be overcome?
% What have others done to address this and what do we propose "instead"
One goal of user augmentation is to increase reaction times, thereby making an interaction more efficient. Previous work on such augmentation scenarios has shown that an actuation preceding a movement by about 80 ms preserves agency~\cite{Kasahara2019-sk}. Evidence from cognitive neuroscience indicates that from around 200ms before a voluntary movement, users are unable to ``veto'' their self-initiated movement~\cite{Schultze-Kraft2016-bx}. Here, the \textit{key} aspect for SoA in action augmentation becomes apparent: External influences on the user's body need to be in line with the user's intention.

In experimental settings, this is typically achieved by using strongly controlled `stimulus-response' paradigms. For example, reaction times when tapping on a touchscreen in response to a presented stimulus on screen, can be reduced by an augmentation device while maintaining SoA for the user~\cite{Kasahara2019-sk, Kasahara2021-gy}. In such controlled scenarios, the timing of the action augmentation can be tuned to be near optimal because the user behavior can be predicted with very high certainty to follow a previously presented stimulus. However, a key challenge remaining is to design systems that maintain agency during unpredictable actions of the user and \textit{without} control over the environment. In other words, how can we design a closed-loop system to deliver a \textit{natural} agency experience for users' augmented actions?

% [4. (short) Answer Teaser
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
% keep this very short]

\subsection{Preserving Agency using Brain Signals of the Intent to (Inter-) act}
% 4. (long) Anwser with teaser image
% Provides the answer on how to overcome the hurdles. Explains how this will help deflect the threats or seize the opportunities.
In this paper, we present a bleeding edge prototype to challenge the disruption in experiencing agency during action augmentation. We developed a brain-computer interface (BCI) that establishes a fast communication channel between a user's brain signals and a physical end effector. The  INTENTION system controls the user's muscles at the time of their \textit{intent to interact}, as measured through readiness potentials (RP) manifesting in the user's electroencephalogram (EEG)~\cite{Nguyen2023-me, Schurger2021-vp, Schultze-Kraft2016-bx, Schultze-Kraft2021-cu}.

% TODO merge this with the figure below
% \missingfigure{maybe small infographic here showing the connection from EEG volitional RP thought to EMS trigger on the Arm. <- better use this for the teaser image. Here better present the main results, i.e. SoA scores? -> Klaus comment: yes, but they don't have to be exclusive. The general setup with the LRP could appear with the main results both in the teaser or here? But the teaser would be more a visualization of the principle, no? That does not necessarily require real results but could show a hyperplane and data dots for examplification.}
% make this as an infographic with comic style drawing of augmented user and then the three pathways drawn
% \begin{figure}
%     \centering
%     \includegraphics[width=\columnwidth]{figures/setup_conditions_draft_small.png}
%     \caption{draft apperatus and conditions}
%     \label{fig:setup}
% \end{figure}
% \begin{figure}[!h]
%     \centering
%     \includegraphics[width=\columnwidth]{figures/augmented_com_model_colour.png}
%     \caption{draft - visualization of comparator model of the SoA in the case of augmented interactions (inspired by Dewey 2019)}
%     \label{fig:com_model}
% \end{figure}

To control an interaction in real-time, the EEG-based RP classifier was used to discriminate between two user states: either the user was predicted to be in an \textit{idle} state, reflecting the absence of an intent to act-- or the user was predicted to be in a \textit{pre-movement} state, indicating the presence of an intent to act. When the system predicted a \textit{pre-movement} state, the user's action was augmented, potentially even before their own voluntary motor command. In our user study, we then applied a mixed-methods research approach to investigate whether keeping the physical impact on the user's body in line with their intention to move preserves their SoA.

%%% Resources
% add results and general outcomes here?

% for now moved to introduction section
% We follow classical work in the neuroscience describing SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay and reproduce it as such. Critically, this temporal compression only occurs following endogenous movement intention. The action outcome is mentally \textit{binded} to the intention. To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay <REF>.

% We found ... rendering the augmentation a \textit{natural}, integrated expansion of the user's body.

% By controlling the stimulus presentation, the experimenter can control the timing of the \textit{pre-emptive} gain of the user's action augmentation.

% Resources
% The authors found a relation between the experience of agency and the \textit{pre-emptive} gain, i.e., how much earlier the action of the user could be elicited. However, setting this gain factor only works in stimulus-response scenarios where the user behavior can be predicted with high accuracy.

% At around 80 ms preceding the naturally occurring action, user's integrate the augmentation and maintain an experience of agency.

% if something happens for you in an interaction there is a cost to that:
% - impacts SoA which in turn may makes user's engage less -> citation?
% - may impact precision and accuracy, which can be very critical in high stakes scenarios, e.g. air traffic control -> citation?

### 2023-09-15 input/3_related_work.tex
pre-submission text

\section{Related Work}
%Intro
In a simplified world, intentional movements directly leading to external events provide a strong sense of responsibility and control over the action's outcome. For instance, when we press a piano key, it is unequivocally clear that we are the ones performing and causing the action. However, the introduction of technology can complicate this relationship. Imagine playing a computer game where a key press controls an avatar playing the piano. The avatar's movement is now generated by a computer, blurring the lines between the performer and the cause of the action. Despite this complexity, research (sources to be provided, such as wn and Imamizu) suggests that as long as the button press and the avatar's movement remain consistent, we still feel a sense of control.

This lingering sense of control might be due to our intentions being reflected in the resulting action (sources needed). However, in augmented interactions, the dynamics change. When technology moves our fingers, our body performs the action, but we may experience a diminished feeling of causing the action [sources - augmented actions lead to less agency; check Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009, Haggard, Clark 2002].

% Why is feeling of control important
Because the feeling of control is related to (check for sources on learning effects etc.), understanding the reason for its loss in augmented interactions is relevant.

% Theorie

The concept of eliciting a sense of agency is commonly explained within the framework of motor control (see Frith and collegus 2000). When we perform an action without any external interference, the sensory consequences of that action align with the predicted sensory consequences, which are based on a copy of motor commands (the efference copy). However, in situations where the movement is elicited by an external cause, as in augmented actions, the predicted sensory consequences may differ or be absent altogether. In such cases, the actual sensory feedback does not match the predicted sensory feedback.

According to this account, the feeling of agency arises only when the predicted sensory feedback perfectly corresponds to the actual sensory feedback. In other words, when our actions and their outcomes align seamlessly, we experience a sense of agency over those actions.

@@

% In this paper, we present a physical action augmentation prototype and a user study to investigate the experience of agency during the use of the prototype. Hence, our research builds on previous qualitative and quantitative research on the sense of agency.
% on actuated haptic systems, brain-computer interfacing as well as
% moving the finger of a user when the user intends to do so, but before the user directs their muscles to move.

%Our research builds on previous engineering work in brain-computer interfacing and on theories of the sense of agency as discussed in cognitive science and neuroscience.

\subsection{The sense of agency in human-computer interaction}
% LT to write first version
\begin{comment}

% Sense of agency in augmented interaction

- intention / voluntary movement is important for sense of agency
- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)

@@

The sense of agency, or this being in the "driving seat when it comes to our own actions"~\cite{Moore2016-ub} can be subdivided into the \textit{feeling} of agency, a low-level pre-reflective sensory process, and the \textit{judgement} of agency, a higher level reflective cognitive process, ~\cite{Moore2016-ub, Danry2022-xk}.

Agency is largely explained with a comparator model, describing internal computational mechanisms of human action control 
-- comparator model (Blakemore 2002, Frith 2000, Frith 2005)
-- integrate predictive coding??

- In the words of Harrard: making an association between action and outcome (Intention - Volition-Movement-Agency -Effect) --> find the source; is from yt video

- Looking at the sense of agency from the perspective of the comparator model
    - if prediction and sensory feedback match = we perceive sense of agency - not only for the big loop with sensory feedback (beep), but also for the forward loop (smaller) --> I can still think that I turned the light on, even if it takes 10 sec after pushing the switch, cause movement still worked as predicted

- In Augmented movements different (explain)
- Before deep diving in these results - excurse on how to measure agency

\subsection{Measuring agency experience}

@@

( van der Wel et al., 2012; Chambon et al., 2013),“ 
Individuals’ sense of agency can be estimated with explicit self-reported measures using Likert’s scales or percentages ( van der Wel et al., 2012; Chambon et al., 2013),“

@@

% - intentional binding: explain what this is and give a bit more detail about the underlying mechanisms and ideas about whats going on in the brain here
- „refers to the perceived temporal attraction between voluntary actions and their outcomes (Haggard et al., 2002). More clearly, the temporal interval between actions and outcomes is perceived as shorter when these outcomes are produced by voluntary actions compared to when they follow, for instance, involuntary movements or external causes (Haggard et al., 2002).“ ([Barlas und Kopp, 2018, p. 2]
- While details about the relationship between measures of intentional binding and SoA are far from clear, they are extensivly used 
- alternatives to intentional binding - sensory attenuation / visual attention

@@

% connection between implicit and explicit measures

- intentional binding effects are sometimes discrepant from explicit judgment of agency (see 30-33 in Wen and Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
- Intentional binding is e.g. influenced by causal relations between events, attention, and arousal (see Wen and Imamizu)

% Sense of agency in augmented interaction

- intention / voluntary movement is important for sense of agency
- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)

% RESEARCH GAP

\subsection{Sense of agency and actuated haptic systems}

\ subsection{Movement Intention}
idea: when we now compare completely voluntary movements with movements triggerd by the brain signal of our intention to move - what happens?

% close this paragraph with the problem that of the control mechanism of the actuator hardware and the disruotion in agency

% PosssessedHand, Gilbert-I miss being me,

@@

>>>>

\section{Related Work}
Our research draws inspiration from neuroscience, especially in assessing the SoA qualitatively and quantitatively, and from engineering work on BCIs as well as on physical user augmentation.

\subsection{Theories of Sense of Agency}
The most widely used theory on how the SoA arises is the \textit{comparator model}~\cite{Blakemore2002-dj, Frith2000-ch, Frith2006-sc}: when we intentionally perform an action, the brain generates sensory predictions about the action outcome. These predictions are constantly compared to the actual sensory data available through the sensory system during the execution of the action. These include continuous signals such as proprioceptive and visual monitoring of the ongoing movement as well as higher level predictions about the semantic outcome of the action~\cite{Clark2013-ah, Haggard2003-ff, Haggard2017-uv}. If no sensorimotor incongruency arises, a SoA manifests.

In the simple case of pressing a key on a piano, the finger movement is constantly compared to the predicted proprioceptive feedback. Subsequently, the tone generated by the key press is evaluated against auditory predictions. On a semantic level, these predictions may be in reference to whether the tone loudness corresponds to the velocity of the key press or whether the tone is in-key or out-of-key~\cite{Pangratz2023-ew}. If these predictions -- based on the intended movement and its expected outcome -- explain the sensory data available, agency is experienced (see figure~\ref{fig:task_design} BASELINE).

In human-computer interaction (HCI) research, these constructs are categorized in their own terms. Often, the term \textit{pre-reflective} is used to describe `early', implicit, experience of agency, such as when matching proprioceptive predictions about finger movements. At higher levels of the cognitive hierarchy, \textit{reflective}, i.e. conscious, experience is used to refer to matching semantic predictions about action outcomes~\cite{Danry2022-xk, Cornelio2022-aq}.

As opposed to intended actions using our own body, movement augmentation hardware allows moving a user's body without their intention. Today, there are three main technologies to physically augment users' actions: Through the use of mechanical actuators, i.e. exoskeletons, a user's body can be moved by applying forces to the extremities. Another possibility is to stimulate the brain directly, so the stimulation causes a motor response, for example by using transcranial magnetic stimulation (TMS). Lastly, electrical muscle stimulation (EMS) makes the user's extremities move by sending current into their muscle activating nerves. In these scenarios, the user computes no predictions about the movement and its outcome. Thus, in reference to the comparator model, none or a decreased SoA arises in the case of externally controlled actions, for example by brain stimulation~\cite{Haggard2002-sz} or when the body is moved by another person~\cite{Kuhn2013-ls}. With something or somebody else moving our body to perform, e.g., a keypress on a piano, the lack of intention to play the piano -- and derived sensory and semantic predictions -- impacts the experience of agency. Such involuntary movements may even cause adverse effects: In case a body movement is elicited by TMS simulation, a resistance to the movement may manifest~\cite{Haggard2002-sz} (see figure~\ref{fig:task_design} NO INTENTION)

\subsubsection{Measuring Sense of Agency}
Both explicit and implicit methods have been developed to evaluate the sense of agency. These methods provide the basis to investigate the effects of action augmentation technology on agency experience. Explicit methods directly query participants to report their subjective experiences using questionnaires. Items such as ``It felt like I was in control of the hand I was looking at''~\cite{Haggard2002-sz} or ``Indicate how much it felt like moving the joystick caused the object on the computer screen to move''~\cite{Ebert2010-lu}, query either the \textit{pre-reflective} action -- or the \textit{reflective} outcome evaluation~\cite{Moore2012-dk}. However, in most cases such questionnaires aim at a higher-level, reflective, judgment of agency.

On the other hand, implicit methods are often used to query low-level pre-reflective sensory predictions that are not consciously reflected~\cite{Moore2016-ub, Limerick2014-un, Moore2012-ic}. Following seminal work in neuroscience has described one effect of SoA as a bias in the perception of action \textit{outcome}: Intentional binding paradigms state that when a button press is followed by a -- delayed -- outcome, participants mentally compress the delay~\cite{Haggard2002-sz}. Critically, this temporal compression only occurs following movement intention. The action outcome is mentally \textit{bound} to the intention. To reduce uncertainty about the binding, the brain `explains away' the excess delta, compressing the action-outcome delay~\cite{Barlas2018-bq}. Supplementing this behavioral phenomenon, physiological evidence can prove useful to further understand the interplay between volitional action and the SoA.

% alternative subsection heading: Brain Signals Reflecting the Intent to (Inter-)Act
% this must make the link clear that we need intention for agency experience and intention is something we can set out to measure using EEG!!!
\subsection{Controlling Actuated Haptic Experiences using Brain Signals Reflecting the Intent to (Inter-)act}
Of the possible physiological signals that can be used to further investigate the experience of agency, the electroencephalogram (EEG) is very well suited because of its high temporal resolution and the non-invasive recording close to the motor command generating structures in the human brain. The Readiness Potential (RP), or \textit{lateralized readiness potential} is an amplitude fluctuation in the ongoing EEG activity that has frequently been observed preceding voluntary action~\cite{Deecke1969-bl, Libet1983-qu}. The RP is reliably observed at electrodes placed over the motor cortex contralateral to the acting hand. Since its measurable onset precedes the time of participants' self-reported conscious movement intention, it has drawn much interest with respect to the debate on free will, see~\cite{Schurger2021-vp} for a recent neuroscientific perspective. However, evidence abounds for its role in action preparation. An RP is typically comprised of two stages: an early slow stage that begins up to two seconds before the actual movement and a late steep stage that starts about 400 milliseconds before movement. The first stage manifests in the pre-supplementary motor area and transfers to the premotor cortex shortly after. The second stage manifests contra-laterally in the primary motor cortex~\cite{Shibasaki2006-mt}.

A recent study has shown that the RP is ingrained in the subconscious mechanisms preceding movements that people cannot explicitly suppress~\cite{Schultze-Kraft2021-cu}. In their study,~\citet{Schultze-Kraft2021-cu} asked participants to find a way to perform voluntary movements while keeping accompanying RP amplitudes as small as possible. After each trial they informed participants about the strength of the RP in the current trial, so participants had a feedback metric to optimize for. They found participants unable to suppress their RP. This inability to suppress the RP renders it a reliable feature for classification. For example, the RP can be detected in real-time using a brain-computer interface (BCI).~\citet{Schultze-Kraft2016-bx} demonstrated a prototype that detects RPs in participants ongoing EEG data and adapts an interface accordingly. Participants were instructed to veto their self-initiated movement whenever a red dot occurred on the screen. The red dot's appearance was controlled by the BCI. Whenever an RP was detected, the red dot appeared. The authors found that participants were able to veto their self-initiated movement if the red dot appeared no later than 200ms preceding their movement onset. After that, participants were unable to ``overwrite'' their motor command and acted regardless of the red dot appearance on screen.

In the present study, we leveraged the RP in a similar way, i.e. to discriminate between two \textit{action} states as the basis to trigger an action augmentation device or not. The technology we used with our prototype, electrical muscle stimulation (EMS), sends currents to the user's forearm leading to a contraction of the flexor muscle moving the ring finger as an augmented movement.

%%% resources

\begin{comment}

@@

@@

@@

@@

% - intentional binding: explain what this is and give a bit more detail about the underlying mechanisms and ideas about whats going on in the brain here
- „refers to the perceived temporal attraction between voluntary actions and their outcomes (Haggard et al., 2002). More clearly, the temporal interval between actions and outcomes is perceived as shorter when these outcomes are produced by voluntary actions compared to when they follow, for instance, involuntary movements or external causes (Haggard et al., 2002).“ ([Barlas und Kopp, 2018, p. 2]
- While details about the relationship between measures of intentional binding and SoA are far from clear, they are extensivly used 
- alternatives to intentional binding - sensory attenuation / visual attention
- involuntary movements produce less binding than do voluntary actions, or even reverse the effect entirely (Yoshi und Harrard 2013, Moore, Wegner und Harrard 2009)
it has been shown, that mere peripheral body movements, elicited by TMS simulations, produce a perceptual repulsion opposite to intentional binding in truly operant intentional actions (Haggard, Clark 2002)

@@

% connection between implicit and explicit measures:
- intentional binding effects are sometimes discrepant from explicit judgment of agency (see 30-33 in Wen and Imamizu) (Saito 2015; Majchrowitz 2018; Ebert
- Intentional binding is e.g. influenced by causal relations between events, attention, and arousal (see Wen and Imamizu)

% Interesting HCI papers:
% PosssessedHand, Gilbert-I miss being me,

% Why is sense of agency important? Why would it be bad if we lost it in augmented humans?
% However, for augmentation hardware to be readily accepted, maintaining a sense of agency is important for several reasons. (find reasons)
% - motor learning ( check Wen, W. et al. Perception and control: individual difference in the sense of agency is associated with learnability in sensorimotor adaptation. Sci. Rep. 11 , 20542 (2021).
% - motivation (find sources)

@@

% This straightforward relationship is complicated by the introduction of technology. Think about, playing a computer game where a key press controls an avatar playing the piano. Here, we do not really press the piano key, the avatar's movement, and the tone are now generated by a computer. Despite the blurring lines between the performer and the cause of the tone, we still feel a sense of authorship when the comparison does not reveal incongruency. This lingering feeling of control is thought to be elicited by our intentions being reflected in the resulting action (sources needed).

% alternative headings:
% Controlling Actuated Haptic experiences using brain signals reflecting the intent to interact
% Triggering EMS using EEG Signals
